---
title: "InClassAssign2"
name: "Jacob Williams"
date: "9/3/2020"
output: html_document
---
Surveys are frequently used to measure political behavior such as voter turnout,
but some researchers are concerned about the accuracy of self-reports. In particular,
they worry about possible social desirability bias where, in postelection surveys,
respondents who did not vote in an election lie about not having voted because they
may feel that they should have voted. Is such a bias present in the American National
Election Studies (ANES)? ANES is a nationwide survey that has been conducted for
every election since 1948. ANES is based on face-to-face interviews with a nationally
representative sample of adults. Table 1.3 displays the names and descriptions of
variables in the turnout.csv data file.

1. Load the data into R and check the dimensions of the data. Also, obtain a
summary of the data. How many observations are there? What is the range of
years covered in this data set?
```{r}
setwd("C:/Users/jwilliamsna/Desktop/BUSH 631/Data")
data_turnout <- read.csv("turnout.csv")

dim(data_turnout)

summary(data_turnout)

("Rows")
nrow(data_turnout)

("Columns")
ncol(data_turnout)
  
VAP <- data_turnout$VAP
("This section is the Voter-Age Population Turnout")
(VAP)

("This dataset encompasses elections from 1980 through 2008, there are over 100 observations in this dataset")

```

2. Calculate the turnout rate based on the voting age population or VAP. Note that
for this data set, we must add the total number of eligible overseas voters since
the VAP variable does not include these individuals in the count. Next, calculate
the turnout rate using the voting eligible population or VEP. What difference do
you observe?
```{r}
('total number of eligible including overseas')
VAP_overseas <- data_turnout$VAP + data_turnout$overseas
(VAP_overseas)

("Turnout Rate including overeas voters")
turnout_VAP_overseas <- (data_turnout$total/ VAP_overseas) * 100

(turnout_VAP_overseas)
("Turnout rate using only the VEP")
turnout_VEP <- (data_turnout$total / data_turnout$VEP) * 100

(turnout_VEP)
("We observe that there is roughly a 2% difference in the turnout rate with the VEP being higher than the VAP+overseas")
```

3. Compute the differences between the VAP and ANES estimates of turnout rate.
How big is the difference on average? What is the range of the differences?
Conduct the same comparison for the VEP and ANES estimates of voter turnout.
Briefly comment on the results.
```{r}
diff <- data_turnout$ANES - turnout_VAP_overseas

(diff)

##(min(diff))
("The range of differences is:")
(max(diff) - min(diff))

diff_VEP <- data_turnout$ANES - turnout_VEP
(diff_VEP)

("This difference shows that the ANES estimate is closer to the voter-eligible population turnout than it is the total voting age population turnout.")


```

4. Compare the VEP turnout rate with the ANES turnout rate separately for
presidential elections and midterm elections. Note that the data set exclude 5. Divide the data into half by election years such that you subset the data into two
periods. Calculate the difference between the VEP turnout rate and the ANES
turnout rate separately for each year within each period. Has the bias of ANES
increased over time?
```{r}
##(turnout_VEP)
names(turnout_VEP) <- data_turnout$year
##(names(turnout_VEP))

presidential_VEP <- subset(turnout_VEP, data_turnout$year %% 4 == 0)##turnout_VEP[seq(from = 1, to = nrow(data_turnout), by = 2)] 


midterm_VEP <- subset(turnout_VEP, data_turnout$year %% 2 == 0 & data_turnout$year %% 4 != 0)##turnout_VEP[seq(from = 2, to = nrow(data_turnout), by = 2)]


turnout_ANES <- data_turnout$ANES
names(turnout_ANES) <- data_turnout$year
##(turnout_ANES)
presidential_ANES <- subset(turnout_ANES, data_turnout$year %% 4 == 0)##turnout_ANES[seq(from = 1, to = nrow(data_turnout), by = 2)]

midterm_ANES <- subset(turnout_ANES, data_turnout$year %% 2 == 0 & data_turnout$year %% 4 != 0)##turnout_ANES[seq(from = 2, to = nrow(data_turnout), by = 2)]

midterm_diff <- midterm_ANES - midterm_VEP
("The difference between ANES and VEP in midterm elections")
(midterm_diff)

presidential_diff <- presidential_ANES - presidential_VEP
("The difference between ANES and VEP in presidential elections")
(presidential_diff)

("In miterm elections years, the ANES has overestimated the turnout while it stayed relatively consistent in presidential elections years.")

```

6. ANES does not interview prisoners and overseas voters. Calculate an adjustment
to the 2008 VAP turnout rate. Begin by subtracting the total number of ineligible
felons and noncitizens from the VAP to calculate an adjusted VAP. Next,
calculate an adjusted VAP turnout rate, taking care to subtract the number of
overseas ballots counted from the total ballots in 2008. Compare the adjusted
VAP turnout with the unadjusted VAP, VEP, and the ANES turnout rate. Briefly
discuss the results.
```{r}
adjusted_VAP <- (data_turnout$VAP) - data_turnout$noncit - data_turnout$felons
names(adjusted_VAP) <- data_turnout$year
(adjusted_VAP['2008'])


adjusted_VAP_turnout <- ((data_turnout$total[14] - data_turnout$overseas[14])/ data_turnout$VAP[14])  * 100##soft code this section

('Adjusted_VAP_turnout')
(adjusted_VAP_turnout)

('Unadjusted VAP')
unadjusted_VAP_turnout <- (data_turnout$total/data_turnout$VAP) * 100 
names(unadjusted_VAP_turnout) <- data_turnout$year
(unadjusted_VAP_turnout['2008'])

("VEP")
names(turnout_VEP) <- data_turnout$year
(turnout_VEP['2008'])

('ANES turnout')
names(turnout_ANES) <- data_turnout$year
(turnout_ANES['2008'])

("The adjusted VAP turnout is lower than every other turnout indicators, which shows that the lack of adjustent for prison and overseas populations will inheritely overexaggerate the rate")
```

